Research Notes

### 2018-02-18 writups/notes.tex
stash

\title{Research Notes}
\author{Yubo Su}

@@

We note that in the no gravity limit that $\rho_1$ does not have an effect on
other dynamical variables, so the equations of motion we must solve are
\begin{equation}
    \begin{split}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} + \frac{\vec{\nabla}P_1}{\rho_0} &= 0.
    \end{split}\label{eq:lin.no_g_eom}
\end{equation}
We can take the divergence of the momentum equation and substitute the
continuity equation to get $\nabla^2 P = 0$. This is a Laplace equation, which
we've solved countless times. Imposing periodic boundary conditions in the $x$
direction and $P_1(z = L) = 0, P_1(z = 0) = \mathcal{P}(x, t)$, we obtain
eigenfunctions
\begin{equation}
    \begin{split}
        P_{1, n}(x, z, t) &=
            \frac{\mathcal{P}_n(t)}{\sinh(k_nL)}e^{ik_nx}\sinh\p*{k_n(L - z)},\\
        u_{1x, n}(x, z, t) &=
            \int\limits^t -\frac{1}{\rho_0}\pd{P_{1, n}}{x}\;\mathrm{d}t,\\
        u_{1z, n}(x, z, t) &=
            \int\limits^t -\frac{1}{\rho_0}\pd{P_{1, n}}{z}\;\mathrm{d}t.
    \end{split}\label{eq:incomp_nog}
\end{equation}
We define $k_n = \frac{2\pi n}{L}, n \in [0, L - 1]$ and $\mathcal{P}(x, t) =
\sum\limits_n \mathcal{P}_n(t)e^{ik_nx}$.

Thus, if we impose BCs $\mathcal{P}(x, t) = \sin \frac{2\pi x}{L}$ and start
with initial conditions such that all quantities are zero, we would expect after
transients die out that
\begin{equation}
    \begin{split}
        P(x, z, t) &= \frac{\sin \frac{2\pi x}{L}}{\sinh 2\pi}
            \sinh \p*{2\pi \frac{L - z}{L}},\\
        u_{1x}(x, z, t) &= -\frac{2\pi t}{L\rho_0}
            \frac{\cos \frac{2\pi x}{L}}{\sinh 2\pi}\sinh \p*{2\pi \frac{L -
                z}{L}},\\
        u_{1z}(x, z, t) &= +\frac{2\pi t}{L\rho_0}
            \frac{\sin \frac{2\pi z}{L}}{\sinh 2\pi}\cosh \p*{2\pi \frac{L -
                z}{L}}.
    \end{split}\label{eq:incomp_nog_sol}
\end{equation}
This is in good agreement with the results, presented in \autoref{fig:no_g}.
Note that $P$ is constant while $\vec{u}$ increases linearly in time, and we
observe the expected $\sim \sin x \sinh \frac{L - z}{z}$ dependence.

@@

    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/prelim/no_g/no_g_P_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/prelim/no_g/no_g_ux_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/prelim/no_g/no_g_uz_t100.png}
    \end{subfigure}
    \caption{$P, u_x, u_z$ at $t = 0.5$ and $t = 1$ for $\rho_0 = 1$.
    }\label{fig:no_g}
\end{figure}

@@

The goal of this will be to lay out a formalism that can reproduce Sutherland et
al.~2011\footnote{DOI:10.1175/JAS-D-11--097.1} and also investigate driven
oscillations versus the breaking of a single wave packet. This represents wave
breaking in the atmosphere.

We adopt notation where $q_0$ is the background quantity and $q_1$ is the
perturbed quantity from the propagating wave.

The fluid equations are
\begin{subequations}\label{se:fl_eq}
    \begin{align}
        \pd{\rho}{t} + \vec{\nabla} \cdot \rho \vec{u} &= 0,
            \label{eq:fl_eq.cont}\\
        \rd{\vec{u}}{t} &= -\vec{\nabla}\frac{P}{\rho} - g\hat{z},
            \label{eq:fl_eq.mom}
    \end{align}
\end{subequations}
where we will take gravity to be uniform throughout the domain of interest. We
will study for no background flow $\vec{u}_0 = 0$ and in the presence of
stratification $\rho_0 \propto e^{-z/H}$. In the absence of any perturbations it
is then easy to show that $\rd{P_0}{z} = -\rho_0 g$.

\subsection{Linear, Incompressible}

\subsubsection{Solution for Arbitrary Stratification}

First, we solve the incompressible case $c_s^2 \to \infty, \vec{\nabla} \cdot
\vec{u} = 0$ in the linear regime. For funsies, we solve for arbitrary
stratification first. The fluid equations to first order reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\rho_0 &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - P_1 \vec{\nabla}\frac{1}{\rho_0}
    \end{split}\label{eq:lin_incomp}
\end{equation}
We expect there to be some $z$ dependence in the amplitude, so we substitute
variables of form $e^{i(kx - \omega t)}$ and do not specify the $z$ dependence.
This gives us
\begin{equation}
    \begin{split}
        -i\omega \rho_1 - u_{1z}\pd{\rho_0}{z} &= 0,\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1}{\rho_0^2}\pd{P_0}{z} &= 0.
    \end{split}
\end{equation}
We substitute $N^2 = -\frac{g}{\rho_0}\pd{\rho_0}{z}$ and $\pd{P_0}{z} = -\rho
g$ to obtain
\begin{subequations}
    \begin{align}
        -i\omega \rho_1 - u_{1z}\frac{\rho_0N^2}{g} &= 0,\label{eq:lin.cont}\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\label{eq:lin.div}\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\label{eq:lin.mom_x}\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1g}{\rho_0} &= 0.\label{eq:lin.mom_z}
    \end{align}
\end{subequations}
Eliminating $u_{1x}$ by substituting~\eqref{eq:lin.div}
into~\eqref{eq:lin.mom_x} and $\rho_1$ by substituting~\eqref{eq:lin.cont}
into~\eqref{eq:lin.mom_z} give
\begin{subequations}
    \begin{align}
        i\omega \pd{u_{1z}}{z} + \frac{k_x^2 P_1}{\rho_0} &= 0,
            \label{eq:lin.eq1}\\
        \p*{\omega^2 - N^2 }u_{1z} + \frac{i\omega}{\rho_0}\pd{P_1}{z} &= 0.
            \label{eq:lin.eq2}
    \end{align}
\end{subequations}
Finally, we multiply~\eqref{eq:lin.eq1} with $\rho_0$ and differentiate
$\mathrm{d}z$ and combine with~\eqref{eq:lin.eq2} to give
\begin{equation}
    \rtd{u_{1z}}{z} + \frac{1}{\rho_0}\pd{\rho_0}{z}\pd{u_{1z}}{z}
        + k_x^2\p*{\frac{N^2}{\omega^2} - 1}u_{1z} = 0.\label{eq:lin.incomp.gen}
\end{equation}

\subsubsection{Introduce Stratification}

With stratification $\rho \propto e^{-z/H}$ \autoref{eq:lin.incomp.gen} clearly
has exponential solutions $e^{\kappa z}$ for
\begin{equation}
    \kappa^2 - \frac{\kappa}{H} + k_x^2\p*{\frac{N^2}{\omega^2} - 1} = 0.
\end{equation}

We permit complex $\kappa = -\frac{1}{2H} + ik_z$, and from the above clearly
\begin{align}
    k_z^2 &= -\frac{1}{4H^2} + k_x^2\p*{\frac{N^2}{\omega^2} - 1},\nonumber\\
    \omega^2 &= \frac{N^2k_x^2}{k_x^2 + k_z^2 + \frac{1}{4H^2}}.
\end{align}

@@

\subsection{Anelastic Solution}

@@

@@

\subsection{Incompressible}

We must bound this to a finite domain. We choose periodic boundary conditions in
$x$ with total length $L_x$, $x \in \s*{-\frac{L}{2}, \frac{L}{2}}$. We choose
$z$ dimension to have total length $L_z$, $z \in [0, L_z]$.

To set up the boundary condition at $z = 0$, we recall that gravity waves in an
atmosphere with $e^{-z/H}$ profile have form
\begin{equation}
    u_{1z} \propto e^{\frac{z}{2H}}e^{i\p*{k_xx + k_zz - \omega t}}.
        \label{eq:bc_below}
\end{equation}
We will take Boussinesq/WKB approximation $H \to 0$, reasonable in the linear
regime, to simplify the BCs. Then $\rho_1 = 0$ (reasonable in the incompressible
limit), $u_{1x} = -\frac{k_z}{k_x}u_{1z}$ and $P_1 = -\frac{\omega
\rho_0}{k_z}u_{1z}$, from the linearized incompressible equations.

The boundary condition at $z = L_z$ is much harder to determine. It is clear
that $\vec{u}_1(z = \infty) = \rho_1(z = \infty) = 0$, and for $L_z \gg H$ this
would be a reasonable approximation, if simply because we expect the majority of
the wave to dissipate via turbulent dissipation as $z$ reaches many $H$. We will
simply choose the BC to be many multiples of $H$. We can solve with both a
Dirichlet and Neumann BC and compare the two solutions; if the solutions
differ significantly then we must choose a larger $L_z$. These are the only two
solutions that can be implemented where we do not need the phase of the linear
wave, which we lose during the nonlinear breaking region, to relate the function
and derivative at the boundary.

@@

We begin our simulation with $\rho_1 = 0, \vec{u}_1 = 0$ strictly within the
domain of simulation. We will borrow some values from Sutherland's paper and use
$k_z = 2\;\mathrm{km^{-1}}$ then define $k_x = -0.4k_z, H = 10 / k_z, A = 0.05 /
k_z, L_z = 300 / k_z, L_x = 20 / k_z$, We also use $\mu \approx 29, T =
273\;\mathrm{K}, \rho_0 = 1\;\mathrm{kg/m^3}, P_0 = \frac{\rho_0 k_BT}{\mu m_p},
g = 10\;\mathrm{m/s^2}$.

>>>>

\title{Research Notes}
\author{Yubo Su}
\rhead{}

@@

We note that in the no gravity limit that $\rho_1$ does not have an effect on
other dynamical variables, so the equations of motion we must solve are
\begin{equation}
    \begin{split}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} + \frac{\vec{\nabla}P_1}{\rho_0} &= 0.
    \end{split}\label{eq:lin.no_g_eom}
\end{equation}
We can take the divergence of the momentum equation and substitute the
continuity equation to get $\nabla^2 P = 0$.

\subsection{Dirichlet BCs}

This is a Laplace equation, which we've solved countless times. Imposing
periodic boundary conditions in the $x$ direction and $P_1(z = L) = 0, P_1(z =
0) = \mathcal{P}(x, t)$, we obtain eigenfunctions
\begin{equation}
    \begin{split}
        P_{1, n}(x, z, t) &=
            \frac{\mathcal{P}_n(t)}{\sinh(k_nL)}e^{ik_nx}\sinh\p*{k_n(L - z)},\\
        u_{1x, n}(x, z, t) &=
            \int\limits^t -\frac{1}{\rho_0}\pd{P_{1, n}}{x}\;\mathrm{d}t,\\
        u_{1z, n}(x, z, t) &=
            \int\limits^t -\frac{1}{\rho_0}\pd{P_{1, n}}{z}\;\mathrm{d}t.
    \end{split}\label{eq:incomp_nog}
\end{equation}
We define $k_n = \frac{2\pi n}{L}, n \geq 0$ and $\mathcal{P}(x, t) =
\sum\limits_n \mathcal{P}_n(t)e^{ik_nx}$.

Thus, if we impose BCs $\mathcal{P}(x, t) = \sin \frac{2\pi x}{L}$ and start
with initial conditions such that all quantities are zero, we would expect after
transients die out that
\begin{equation}
    \begin{split}
        P(x, z, t) &= \frac{\sin \frac{2\pi x}{L}}{\sinh 2\pi}
            \sinh \p*{2\pi \frac{L - z}{L}},\\
        u_{1x}(x, z, t) &= -\frac{2\pi t}{L\rho_0}
            \frac{\cos \frac{2\pi x}{L}}{\sinh 2\pi}\sinh \p*{2\pi \frac{L -
                z}{L}},\\
        u_{1z}(x, z, t) &= +\frac{2\pi t}{L\rho_0}
            \frac{\sin \frac{2\pi z}{L}}{\sinh 2\pi}\cosh \p*{2\pi \frac{L -
                z}{L}}.
    \end{split}\label{eq:incomp_nog_sol}
\end{equation}
This is in good agreement with the results, presented in \autoref{fig:no_g}.
Note that $P$ is constant while $\vec{u}$ increases linearly in time, and we
observe the expected $\sim \sin x \sinh \frac{L - z}{z}$ dependence. In fact,
$u_{1x}, u_{1z}$ are exactly $\frac{2\pi}{10}$ at $t = 1$.

@@

    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/prelim/no_g/no_g_P_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/prelim/no_g/no_g_ux_t100.png}
    \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../sims/prelim/no_g/no_g_uz_t100.png}
    \end{subfigure}
    \caption{$P, u_x, u_z$ at $t = 0.5$ and $t = 1$ for $\rho_0 = 1$. We choose
    a $L = 10$ square domain.}\label{fig:no_g}
\end{figure}

@@

\subsection{Sommerfield/Radiative + Driving BCs}

This is the more interesting case. Let's go back to the Laplace equation
$\nabla^2 P = 0$ but instead implement a driving term at $z = 0$ and radiative
BCs at $z = L$. This

\section{Incompressible, Stratified w/ Gravity}

\subsection{Eigenfunctions}

Let's restore the $\rho_1 g$ term now. For funsies, we begin by solving for
arbitrary stratification $\rho_0(z)$ first. The fluid equations to first order
reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\rho_0 &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - \frac{\rho_1 g}{\rho_0}
    \end{split}\label{eq:lin_incomp}
\end{equation}
We expect there to be some $z$ dependence in the amplitude, so we substitute
variables of form $e^{i(kx - \omega t)}$ and do not specify the $z$ dependence.
This gives us
\begin{equation}
    \begin{split}
        -i\omega \rho_1 - u_{1z}\pd{\rho_0}{z} &= 0,\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1 g}{\rho_g} &= 0.
    \end{split}
\end{equation}
We substitute $N^2 = -\frac{g}{\rho_0}\pd{\rho_0}{z}$ to obtain
\begin{subequations}
    \begin{align}
        -i\omega \rho_1 - u_{1z}\frac{\rho_0N^2}{g} &= 0,\label{eq:lin.cont}\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\label{eq:lin.div}\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\label{eq:lin.mom_x}\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1g}{\rho_0} &= 0.\label{eq:lin.mom_z}
    \end{align}
\end{subequations}
Eliminating $u_{1x}$ by substituting~\eqref{eq:lin.div}
into~\eqref{eq:lin.mom_x} and $\rho_1$ by substituting~\eqref{eq:lin.cont}
into~\eqref{eq:lin.mom_z} give
\begin{subequations}
    \begin{align}
        i\omega \pd{u_{1z}}{z} + \frac{k_x^2 P_1}{\rho_0} &= 0,
            \label{eq:lin.eq1}\\
        \p*{\omega^2 - N^2 }u_{1z} + \frac{i\omega}{\rho_0}\pd{P_1}{z} &= 0.
            \label{eq:lin.eq2}
    \end{align}
\end{subequations}
Finally, we multiply~\eqref{eq:lin.eq1} with $\rho_0$ and differentiate
$\mathrm{d}z$ and combine with~\eqref{eq:lin.eq2} to give
\begin{equation}
    \rtd{u_{1z}}{z} + \frac{1}{\rho_0}\pd{\rho_0}{z}\pd{u_{1z}}{z}
        + k_x^2\p*{\frac{N^2}{\omega^2} - 1}u_{1z} = 0.\label{eq:lin.incomp.gen}
\end{equation}

Let's now pick stratification $\rho \propto e^{-z/H}$
\autoref{eq:lin.incomp.gen} clearly has exponential solutions $e^{\kappa z}$ for
\begin{equation}
    \kappa^2 - \frac{\kappa}{H} + k_x^2\p*{\frac{N^2}{\omega^2} - 1} = 0.
\end{equation}
We permit complex $\kappa = \frac{1}{2H} + ik_z$, and from the above clearly
\begin{align}
    k_z^2 &= -\frac{1}{4H^2} + k_x^2\p*{\frac{N^2}{\omega^2} - 1},\nonumber\\
    \omega^2 &= \frac{N^2k_x^2}{k_x^2 + k_z^2 + \frac{1}{4H^2}}.
\end{align}
Thus the eigenfunctions are
\begin{equation}
    \begin{split}
        u_{1z} &= e^{z/2H} e^{i(k_zz + k_xx - \omega t)},\\
        u_{1x} &= -\frac{k_z + i/2H}{k_x} u_{1z},\\
        \rho_1 &= \frac{i \rho_0}{H\omega} u_{1z},\\
        P_1 &= -\frac{\rho_0 \omega}{k_x^2}\p*{k_z + i/2H}u_{1z}.
    \end{split}\label{eq:strat_eigens}
\end{equation}

\subsection{Solving an IVP}

\section{(Algebra) The Anelastic/Boussinesq Approximations}

@@

\subsection{Anelastic Solution to Stratified Atmosphere}

@@

TODO (fluid equations, driven on bottom, parameters)

@@

TODO (periodic in x, show that right number of BCs in $z$, discuss whether gauge
choice).

@@

TODO (CFL conditions etc.)
